---
title: "Home Credit Modeling"
format: html
---

# Modeling Workflow

Complete modeling pipeline for Home Credit default prediction.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(caret)
library(randomForest)
library(pROC)
library(ROSE)

calc_auc <- function(truth, estimate) {
  auc_obj <- pROC::roc(as.numeric(truth), as.numeric(estimate), quiet = TRUE)
  as.numeric(auc_obj$auc)
}

impute_numeric_median <- function(df) {
  df %>% mutate(across(where(is.numeric), ~ if_else(is.na(.), median(., na.rm = TRUE), .)))
}

app_train_final <- read_csv("application_train_final.csv", show_col_types = FALSE)
app_test_final <- read_csv("application_test_final.csv", show_col_types = FALSE)

train <- app_train_final

demo_cols <- c(
  "DAYS_BIRTH", "DAYS_EMPLOYED", "DAYS_ID_PUBLISH", "DAYS_LAST_PHONE_CHANGE",
  "AGE_YEARS", "EMPLOYMENT_YEARS", "CNT_FAM_MEMBERS"
)
demo_cols <- demo_cols[demo_cols %in% names(train)]

cat("Data loaded. Train:", nrow(app_train_final), "rows. Test:", nrow(app_test_final), "rows.\n")
```

# 1. Baseline

Establish a majority-class and simple EXT_SOURCE benchmark.

```{r baseline, message=FALSE, warning=FALSE}
majority_acc <- mean(train$TARGET == 0)
cat("Majority class accuracy:", round(majority_acc, 4), "\n")

ext_cols <- c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")
ext_cols <- ext_cols[ext_cols %in% names(train)]

if (length(ext_cols) > 0) {
  train_simple <- train %>% select(TARGET, all_of(ext_cols)) %>% drop_na()
  train_simple$score <- rowMeans(as.matrix(train_simple %>% select(all_of(ext_cols))), na.rm = TRUE)
  baseline_auc <- calc_auc(train_simple$TARGET, train_simple$score)
  cat("EXT_SOURCE baseline AUC:", round(baseline_auc, 4), "\n")
}
```

# 2. Model Comparison

Compare Logistic Regression, Naive Bayes, and Random Forest with 3-fold CV using AUC.

```{r model_comparison, message=FALSE, warning=FALSE}
library(naivebayes)

numeric_cols <- train %>% select(where(is.numeric)) %>% select(-SK_ID_CURR, -TARGET) %>% names()
model_data <- train %>% select(TARGET, all_of(numeric_cols)) %>% drop_na()

set.seed(42)
folds <- createFolds(model_data$TARGET, k = 3, list = TRUE)

ext_source_cols <- c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")
ext_source_cols <- ext_source_cols[ext_source_cols %in% names(model_data)]
combined_cols <- unique(c(demo_cols, ext_source_cols))

cv_results <- tibble(Model = character(), Fold = integer(), AUC = double())

for (i in seq_along(folds)) {
  idx <- folds[[i]]
  tr <- model_data[-idx, ]
  te <- model_data[idx, ]

  fit_lr <- glm(reformulate(demo_cols, "TARGET"), data = tr, family = binomial)
  pr_lr <- predict(fit_lr, newdata = te, type = "response")
  cv_results <- cv_results %>% add_row(Model = "LR_Demo", Fold = i, AUC = calc_auc(te$TARGET, pr_lr))

  tr_nb <- tr %>% mutate(TARGET = factor(TARGET, levels = c(0, 1)))
  fit_nb <- naive_bayes(reformulate(combined_cols, "TARGET"), data = tr_nb)
  pr_nb <- predict(fit_nb, newdata = te, type = "prob")[, "1"]
  cv_results <- cv_results %>% add_row(Model = "NaiveBayes", Fold = i, AUC = calc_auc(te$TARGET, pr_nb))

  tr_rf <- tr %>% mutate(TARGET = factor(TARGET, levels = c(0, 1)))
  fit_rf <- randomForest(reformulate(combined_cols, "TARGET"), data = tr_rf, ntree = 100)
  pr_rf <- predict(fit_rf, newdata = te, type = "prob")[, "1"]
  cv_results <- cv_results %>% add_row(Model = "RandomForest", Fold = i, AUC = calc_auc(te$TARGET, pr_rf))
}

cv_results %>%
  group_by(Model) %>%
  summarise(Mean_AUC = mean(AUC), SD_AUC = sd(AUC), .groups = "drop") %>%
  arrange(desc(Mean_AUC))
```

# 3. Class Imbalance

Evaluate downsampling, upsampling, and SMOTE for LR on demographic features.

```{r imbalance, message=FALSE, warning=FALSE}
set.seed(42)
down <- ovun.sample(TARGET ~ ., data = model_data, method = "under")$data
up <- ovun.sample(TARGET ~ ., data = model_data, method = "over")$data
smote <- ROSE::ROSE(TARGET ~ ., data = model_data, seed = 42)$data

cv_lr_auc <- function(df) {
  flds <- createFolds(df$TARGET, k = 3, list = TRUE)
  aucs <- map_dbl(flds, function(idx) {
    tr <- df[-idx, ]
    te <- df[idx, ]
    fit <- glm(reformulate(demo_cols, "TARGET"), data = tr, family = binomial)
    pr <- predict(fit, newdata = te, type = "response")
    calc_auc(te$TARGET, pr)
  })
  mean(aucs)
}

tibble(
  Strategy = c("Original", "Downsample", "Upsample", "SMOTE"),
  Mean_AUC = c(cv_lr_auc(model_data), cv_lr_auc(down), cv_lr_auc(up), cv_lr_auc(smote))
)
```

# 4. Hyperparameter Tuning

Use randomized search (~20 iterations) on a 5k subsample with 3-fold CV for Random Forest.

```{r tuning, message=FALSE, warning=FALSE}
set.seed(42)
train_up <- up

tune_df <- train_up %>%
  select(TARGET, all_of(combined_cols)) %>%
  impute_numeric_median() %>%
  sample_n(min(5000, n()))

flds <- createFolds(tune_df$TARGET, k = 3, list = TRUE)

param_grid <- tibble(
  mtry = sample(1:max(1, length(combined_cols)), 20, replace = TRUE),
  ntree = sample(c(100, 150, 200, 300), 20, replace = TRUE),
  nodesize = sample(c(1, 5, 10, 20), 20, replace = TRUE)
)

tuning_results <- tibble(iter = integer(), mtry = integer(), ntree = integer(), nodesize = integer(), Mean_AUC = double())

for (j in seq_len(nrow(param_grid))) {
  p <- param_grid[j, ]
  aucs <- map_dbl(flds, function(idx) {
    tr <- tune_df[-idx, ] %>% mutate(TARGET = factor(TARGET, levels = c(0, 1)))
    te <- tune_df[idx, ]
    fit <- randomForest(reformulate(combined_cols, "TARGET"), data = tr, mtry = p$mtry, ntree = p$ntree, nodesize = p$nodesize)
    pr <- predict(fit, newdata = te, type = "prob")[, "1"]
    calc_auc(te$TARGET, pr)
  })

  tuning_results <- tuning_results %>%
    add_row(iter = j, mtry = p$mtry, ntree = p$ntree, nodesize = p$nodesize, Mean_AUC = mean(aucs, na.rm = TRUE))
}

best_params <- tuning_results %>% arrange(desc(Mean_AUC)) %>% slice(1)
best_params
```

# 5. Supplementary Features

Use aggregated supplementary features and compare with/without them.

```{r supplementary_features, message=FALSE, warning=FALSE}
supp_prefix <- c("BUREAU_", "PREV_APP_", "INSTALL_")
supp_cols <- names(train)[str_detect(names(train), paste(supp_prefix, collapse = "|"))]
supp_cols <- supp_cols[vapply(train[supp_cols], is.numeric, logical(1))]

base_features <- demo_cols
enriched_features <- unique(c(base_features, supp_cols))

compare_rf_auc <- function(feature_set) {
  df <- train %>%
    select(TARGET, all_of(feature_set)) %>%
    impute_numeric_median()

  flds <- createFolds(df$TARGET, k = 3, list = TRUE)

  aucs <- map_dbl(flds, function(idx) {
    tr <- df[-idx, ] %>% mutate(TARGET = factor(TARGET, levels = c(0, 1)))
    te <- df[idx, ]
    fit <- randomForest(
      reformulate(feature_set, "TARGET"),
      data = tr,
      ntree = best_params$ntree,
      mtry = min(best_params$mtry, length(feature_set)),
      nodesize = best_params$nodesize
    )
    pr <- predict(fit, newdata = te, type = "prob")[, "1"]
    calc_auc(te$TARGET, pr)
  })

  mean(aucs)
}

auc_base <- compare_rf_auc(base_features)
auc_enriched <- compare_rf_auc(enriched_features)

supp_results <- tibble(
  Model = c("RF (Demographics)", "RF (Demographics + Supplementary)"),
  Mean_AUC = c(auc_base, auc_enriched)
)

supp_results

final_features <- if (auc_enriched >= auc_base) enriched_features else base_features
```

# 6. Final Submission

Predict default probabilities for each SK_ID_CURR in test and save `submission.csv`.

```{r submission, message=FALSE, warning=FALSE}
final_train <- train %>%
  select(TARGET, all_of(final_features)) %>%
  impute_numeric_median() %>%
  mutate(TARGET = factor(TARGET, levels = c(0, 1)))

rf_final <- randomForest(
  reformulate(final_features, "TARGET"),
  data = final_train,
  ntree = best_params$ntree,
  mtry = min(best_params$mtry, length(final_features)),
  nodesize = best_params$nodesize
)

test_pred_df <- app_test_final %>%
  select(SK_ID_CURR, all_of(final_features)) %>%
  impute_numeric_median()

probs <- predict(rf_final, newdata = test_pred_df, type = "prob")[, "1"]

submission <- tibble(
  SK_ID_CURR = test_pred_df$SK_ID_CURR,
  TARGET = as.numeric(probs)
)

readr::write_csv(submission, "submission.csv")
cat("Wrote submission.csv with", nrow(submission), "rows\n")
```

---
End of modeling notebook.

The random forest model got a score of 0.66 on the Kaggle competition.
